\documentclass[10pt]{article}
\usepackage[OT1]{fontenc}
\newtheorem{define}{Definition}
\usepackage{mathtools}
\newcommand{\R}{\mathbf{R}}
\newcommand{\transpose}{T}
\newcommand{\diag}{\operatorname{diag}}
\newcommand{\vol}{\operatorname{vol}}
\newcommand{\conv}{\operatorname{conv}}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{hyperref}
\usepackage[style=alphabetic, backend=biber]{biblatex}
\bibliography{refs.bib}

\oddsidemargin=0.15in
\evensidemargin=0.15in
\topmargin=-.5in
\textheight=9in
\textwidth=6.25in

\begin{document}
\input{preamble.tex}
\lecture{11}{April 27, 2015}{Ola Svensson}{Louis Faucon, Andreas Haupt}

\section{Introduction}
In today's lecture, the \emph{Ellipsoid Method} is introduced, which has been the first weakly polynomial-time algorithm\footnote{An algorithm is weakly polynomial if its running time not only depends on the number of inputs, but also logarithmically on their size.} for the solution of LPs. The method was originally proposed in \cite{shor77} for solving convex optimization problems. In \cite{khachiyan79}, this technique was applied to Linear Programming. The method is fast in theory, but very slow in practice as experiments show. A major advantage of the Ellipsoid method is that it can solve even exponentially-sized LPs for which an efficient so-called separation oracle exists. This is the case i.e. for the Traveling Salesperson problem, as we show in \autoref{subsec:seporacle}. Another widely used solution method for LPs, the \emph{Interior Point Method} introduced by \cite{karmarkar89}, will not be covered here, as it does not have this advantage, although it is proved efficient in theory and practice, but which will not be covered here. 

\subsection{Motivation}
Following \cite{lovaszschrijver}, we consider the problem of catching a lion in Sahara with the assumption, that there is at most one lion. This can be done by building a fence around Sahara, and iteratively halving Sahara or already halved parts of it by a fence and find out anyhow, on which side of this halving fence the lion is. When the area is smaller than the smallest lion, we catch it. Questions arise :
\begin{itemize}
\item How big is Sahara?
\item When to terminate, i.e. How small is the smallest lion?
\end{itemize}

\subsection{Problem Statement}
The problem we wish to solve is the following:

Given a bounded convex set $P\subseteq \R^n$, find: $x \in P$ or decide, that $P = \emptyset$.

This problem seems to solve only feasibility, but since we can add the cost function as as one of the limiting inequalities the algorithm remains very useful for optimization.

\section{The Ellipsoid Algorithm}
The following routine solves the Feasibility Problem:
\begin{enumerate}
\item Start with an Ellipsoid $E_0 \supseteq P$. 
\item While the center of $a_k$ of $E_k$ is not in $P$ select $c \in \R^n$, such that
\[
\{x \in \R^n | c^\transpose x < c^\transpose a_k \} \supseteq P
\]
i.e. pick a hyperplane that \emph{separates} $P$.

and pick $E_{k+1}$ the smallest possible ellipsoid, such that
\[
E_{k+1} \supseteq E_k \cap \{x \in \R^n | c^\transpose x < c^\transpose a_k \}
\]
\end{enumerate}

\subsection{Ellipsoids}
Recall that $A \in \R^{n \times n}$ is called positive definite, if it is symmetric, i.e. $A = A^\transpose$ and 
\[
\forall x \in \R^n\setminus \{0\} \colon x^\transpose A x > 0
\]
i.e. that the scalar product induced by $A$ is positive definite. This class of matrices is closed under inversion and for any positive definite matrix $A$ there exists a matrix $A^{\frac{1}{2}}$, such that
\[
A = A^{\frac{1}{2}} (A^{\frac{1}{2}})^\transpose
\]
\begin{definition}[Ellipsoid]
An ellipsoid $E(a,A)$ with center $a \in \R^n$ and positive definite matrix $A$ is defined as
\[
E(a,A) \coloneqq \{x \in \R^n | (x-a)^{\transpose} A^{-1} (x-a) \le 1 \} = \{x \in \R^n | \| (A^{-1})^\frac{1}{2} (x-a) \| \le 1 \}
\]
\end{definition}
In fact, $E(a,A)$ is just the image of the unit ball $\{x\in \R^n | \|x\|\le 1\}$ under the affine-linear map $x \mapsto A^{\frac{1}{2}} x + a$. As an example, consider $E(a,A)$ with 
\begin{align}
A^{\frac{1}{2}} &= \begin{pmatrix} 2 & 0 \\ 0 & 1 \end{pmatrix} & a &= \begin{pmatrix} 4 \\ 0 \end{pmatrix}. \label{eq:ellipsoidexample}
\end{align}
The result is \autoref{fig:ellipsoidexample}.
\begin{figure}
\centering
\begin{tikzpicture}[scale=2]
\begin{scope}
    \draw[->] (-1.1,0) -- (6.1,0) node[right] {$x$} coordinate(x axis);
    \draw[->] (0,-1.1) -- (0,1.1) node[above] {$y$} coordinate(y axis);
    \foreach \x/\xtext in {-1,0,...,6}
      \draw[xshift=\x cm] (0pt,1pt) -- (0pt,-1pt) node[below,fill=white] {$\xtext$};
    \foreach \y/\ytext in {-1, -.5/-\frac{1}{2}, .5/\frac{1}{2}, 1}
      \draw[yshift=\y cm] (1pt,0pt) -- (-1pt,0pt) node[left,fill=white] {$\ytext$};
  \end{scope}
\draw (0,0) circle (1cm);
\draw (4,0) ellipse (2cm and 1cm);
\path (.4,1) edge [bend left, ->] node[above]{$x \mapsto \begin{pmatrix} 2 & 0 \\ 0 & 1 \end{pmatrix} x +  \begin{pmatrix} 4 \\ 0 \end{pmatrix}$} (3.4,1);
\end{tikzpicture}
\caption{On the left the unit ball in $\R^2$, the affine map mapping the unit circle to the ellipsoid from \protect\autoref{eq:ellipsoidexample}.}
\label{fig:ellipsoidexample}
\end{figure}
\begin{lemma}[Half-Ball Lemma]\label{lem:halfball}
The half-ball
\[
H = \{x \in \R^n | \| x \| \le 1, x_1 \ge 0 \}
\]
is contained in the ellipsoid
\[
E = \left\{ x \in \R^n \middle| \left(\frac{n+1}{n}\right)^2\left(x_1 - \frac{1}{n+1}\right)^2 + \frac{n^2-1}{n^2} \sum_{i=2}^n x_i^2 \le 1 \right\}
\]
\end{lemma}
\begin{proof}
Let $x$ be in the half-ball, we have $\sum_{i=1}^n x_i^2 \le 1, x_1 \ge 0$, then one calculates :
\begin{align*}
\left(\frac{n+1}{n} \right)^2 \left(x_1-\frac{1}{n+1}\right)^2 + \frac{n^2-1}{n^2} \sum_{i=2}^n x_i^2 
&\leq \left(\frac{n+1}{n} \right)^2 \left(x_1-\frac{1}{n+1}\right)^2 + \frac{n^2-1}{n^2}(1 - x_1^2)\\
&\leq \frac{2(n+1)}{n^2} x_1^2-\frac{2(n+1)}{n^2}x_1 + 1\\
&\leq 1 \text{ because } 0 \leq x_1 \leq 1
\end{align*}
\end{proof}


We remark, that this ellipsoid is $E(a,A)$ with
\begin{align*}
a &= \frac{1}{n+1} e_1 &  A^{\frac{1}{2}}&= \diag \left(\frac{n}{n+1}, \frac{n}{\sqrt{n^2-1}}, \dots, \frac{n}{\sqrt{n^2-1}}\right),
\end{align*}
where $e_1$ is the first canonical basis vector and $\diag (x_1, \dots, x_n)$ denotes the matrix with $x_1, \dots, x_n$ on its diagonal. In dimension two, this ellipsoid is given by \autoref{fig:halfball}.
\begin{figure}
\centering
\begin{tikzpicture}[scale=2]
\begin{scope}
    \draw[->] (-1.6,0) -- (1.6,0) node[right] {$x$} coordinate(x axis);
    \draw[->] (0,-1.1) -- (0,1.1) node[above] {$y$} coordinate(y axis);
    \foreach \x/\xtext in {-1,0,...,1}
      \draw[xshift=\x cm] (0pt,1pt) -- (0pt,-1pt) node[below,fill=white] {$\xtext$};
    \foreach \y/\ytext in {-1, -.5/-\frac{1}{2}, .5/\frac{1}{2}, 1}
      \draw[yshift=\y cm] (1pt,0pt) -- (-1pt,0pt) node[left,fill=white] {$\ytext$};
  \end{scope}
\draw (1,0) arc (0:180:1cm);
\draw (0,0.33333) ellipse (1.15 and 0.6666);
\end{tikzpicture}
\caption{The smallest ellipsoid containing the half-ball in dimension two}
\label{fig:halfball}
\end{figure}
What is the volume of our ellipsoid? For sure, $a$ does not matter for the volume. The change of volume is thus given by $\det (A^{\frac{1}{2}})$ and we have
\[
\det (A^{\frac{1}{2}}) = \frac{n}{n+1} (\frac{n}{\sqrt{n^2-1}})^{n-1} \le e^{-\frac{1}{n+1}} (e^{\frac{1}{n^2-1}})^{\frac{n-1}{2}} = e^{-\frac{1}{2(n+1)}}
\]

using $e^x \geq 1 + x$.
\begin{corollary}
If $E_n$ denotes the unit ball in $n$ dimensions, and $E$ the corresponding ellipsoid from Lemma \ref{lem:halfball}, then
\[
\frac{\vol (E)}{\vol (E_n)} = \det (A^{\frac{1}{2}}) \le e^{-\frac{1}{2(n+1)}}<1
\]
\end{corollary}
More generally, we may consider the half-ball in any direction $d$. Then the ellipsoid becomes $E = (-\frac{d}{n+1}, F)$, where $F = \frac{n^2}{n^2-1} (I - \frac{2}{n+1} dd^\transpose)$, where $I$ denotes the identity matrix. Now, given $E_k = (a_k, A_k)$ and some hyperplane which goes through $a_k$, we can apply the inverse transformation $T^{-1} (x) = (A^{-1})^{\frac{1}{2}} (x-a)$ :
\begin{itemize}
\item $E_k$ becomes the unit ball. 
\item The hyperplane now cuts the unit ball in half-balls $\{x\in \R^n | c^\transpose x \le c^\transpose a_k \} \xmapsto{T^{-1}} \{y\in \R^n | c^\transpose (A^{\frac{1}{2}} y + a_k ) \le c^\transpose a_k\} = \{y \in \R^n | (c^\transpose A^{\frac{1}{2}}) y \le 0 \}$
\end{itemize}
Finally applying again the transformation to the ellipsoid given by Lemma \ref{lem:halfball} gives us the desired ellipsoid.


\section{Stopping Criterion, Minimal Size of Polytopes and choice of separation hyperplane for binary problems}
We now discuss three remaining questions:
\begin{enumerate}
\item What is the minimal volume of a non-empty polytope?
\item How to select $E_0$?
\item How to find the separation $c$?
\end{enumerate}
In the following two subsections, we target the first two questions and the third is dealt with in section \ref{sec:oracles}.
\subsection{Minimum Size of a polytope}
We are interested in the case, when $P = \conv (S), S \subseteq \{0,1\}^n$. We assume, that $P$ is full-dimensional. What is our starting ellipsoid? We may take $a = (\frac{1}{2}, \dots, \frac{1}{2})$, with radius $\sqrt{n}$, which has volume smaller than $\sqrt{n}^{n} 2^n$, which means $\log (\vol (E_0)) = O(n \log n)$. 
\subsection{Stopping Criterion}
We can quit, when 
\begin{align*}
\vol (E_k) &< \vol(P) \\
\vol (E_k) &\le e^{-\frac{k}{2(n+1)}} \vol (E_0)
\end{align*}
This yields, that at most $O(n \log (\frac{\vol (E_0)}{\vol (P)}))$%:
iterations.

Let $v_0 \in P$. Since $P$ is full-dimensional, there exist $v_1, \dots, v_n \in P$, such that the convex hall of this point set is a simplex. It has volume
\[
\frac{1}{n!} \left\lvert\det ((v_0-v_1, \dots, v_0-v_n))\right\rvert
\]
Therefore, $\vol (P) \ge \frac{1}{n!}$, which means we have a polynomial number of iterations.

\section{Separation Oracles}
\label{sec:oracles}
The running time of the Ellipsoid Method is independent of the number of constraints. To solve the feasibility problem for a convex $P$ with the Ellipsoid Method, you only need to design a separation oracle.

Separation Oracle : 
Given a point $x^*$,
decide whether $x^*\in P$ and if not output $a \in \R^n, b \in \R$, such that $a^\transpose x \le b, \forall x \in P$ and $a^\transpose x^* > b$.



\subsection{Separation Oracle for Travelling Salesperson Problem}\label{subsec:seporacle}
The Travelling Salesperson Problem asks for a minimum length Hamiltonian circuit in an undirected weighted graph $G$. For $x_e \in \{0,1\}^{E(G)}$ we let $x_e$ be the incidence vector of a tour. A popular LP relaxation for this problem is
\begin{alignat}{3}
\min &\sum_{e \in E(G)} d(e) x_e&&\nonumber \\
&\sum_{e \in \delta (v)} x_e &=2 & \forall v \in V(G) \label{eq:cutinequalities}\\
&\sum_{e \in \delta(S)} x_e \ge 2 & \forall S \subseteq V\nonumber\\
& x &\ge 0\nonumber
\end{alignat}
This LP can be solved efficiently using the Ellipsoid method. The inequalities \autoref{eq:cutinequalities} can be efficiently separated by calculating the minimum $s$-$t$-cut for all $s,t \in V$. 


\printbibliography

\end{document}